---
title: "Cencored regression, weak IV, and quantile regression"
subtitle: "Tutorial 1"
date: "Stanislav Avdeev"
output:
  xaringan::moon_reader:
    self_contained: TRUE
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    # Run xaringan::summon_remark() for this
    #chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE) 
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(gganimate)
library(estimatr)
library(magick)
library(dagitty)
library(ggthemes)
library(directlabels)
library(ggdag)
library(fixest)
library(jtools)
library(scales)
library(Cairo)
library(fabricatr)
theme_metro <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_void_metro <- function(x) {
  theme_void() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
theme_metro_regtitle <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```


# Tutorials

* 7 TA sessions
* 6 TA sessions are about lecture material
* The last session is primarily about exam and remaining questions about the course material (TBA)
* Send me any questions you want to discuss before each TA session
* Use Canvas or send me an email
* Alternately, leave your questions anonymously here: https://www.menti.com/c6uyd9qan4 (I will update the link each week on Canvas)

---

# Assignments

* Due date: 11:59pm on Sundays (the first assignment is an exception: 11:59am on Tuesday)
* Assignments are graded within a week from the deadline
* Solutions will not be shared so if you want to discuss a specific exercise, let me know before the TA session (you submit your solutions on Sunday, thus, we can discuss any questions on the following TA session on Tuesday)

---

# Course objective
* The key objective of the course is **applying** microeconometric techniques rather than **deriving** econometric and statistical properties of estimators.
* In other words, there’s way less of this:

\begin{align*}
  \text{plim} \hat{\beta}_{OLS} = \beta + \text{plim} (\frac{1}{N}X'X)^{-1} \text{plim} \frac{1}{N} X' \epsilon = \beta + Q^{-1} \times \text{plim} \frac{1}{N} X' \epsilon
\end{align*}

* And way more of this:

```{r, echo=TRUE, eval=FALSE}
library(fixest)

tb <- tibble(groups = sort(rep(1:10, 600)), time = rep(sort(rep(1:6, 100)), 10)) %>%
  mutate(Treated = I(groups > 5) * I(time > 3)) %>%
  mutate(Y = groups + time + Treated*5 + rnorm(6000))
m <- feols(Y ~ Treated | groups + time, data = tb)
```

If you would like to go deeper into the former, take Advanced Econometrics I and II next year

---

# Weak instrument problem

- Weak instrument problem means that we probably shouldn't be using IV in small samples
- This also means that it's really important that $cov(X, Z)$ is not small
- If $Z$ has only a trivial effect on $X$, then it's not *relevant* - even if it's truly exogenous, it does not matter because there's no variation in $X$ we can isolate with it
- And our small-sample bias will be big

---

# Weak instrument problem

- There are some rules of thumb for how strong an instrument must be to be counted as "not weak"
- A t-statistic above 3, or an F statistic from a joint test of the instruments that is 10 or above
- These rules of thumb aren't great - selecting a model on the basis of significance naturally biases your results
- What you really want is to know the *population* effect of $Z$ on $X$ - you want the F-statistic from *that* to be 10+. Of course we don't actually know that.

---

# Weak instrument problem: simulation

- Let's look at the output of `feols()` using a simulated dataset

```{r, echo = TRUE}
library(fabricate)
set.seed(777)

df <- fabricate(
  N = 200,
  Y = rpois(N, lambda = 4),
  Z = rbinom(N, 1, prob = 0.4),
  X1 = Z * rbinom(N, 1, prob = 0.8),
  X2 = rnorm(N),
  G = sample(letters[1:4], N, replace = TRUE)
)
```

---

# Weak instrument problem: simulation

```{r, echo = TRUE}
iv <- feols(Y ~ X2 | X1 ~ Z, data = df, se = 'hetero')
thef <- fitstat(iv, 'ivf', verbose = FALSE)$`ivf1::X1`$stat
iv
```

- `r scales::number(thef, accuracy = .01)` is way above 10! We're probably fine in this particular regression

---

# Overidentification tests

- "Overidentification" just means we have more identifying conditions (validity assumptions) than we actually need. We only need one instrument, but we have two (or more)
- So we can compare what we get using each instrument individually
- If we assume that *at least one of them is valid*, and they both produce similar results, then that's evidence that *both* are valid

---

# Overidentification tests: simulation

- We can do this using `diagnostics = TRUE` in `iv_robust` again

```{r, echo = TRUE}
set.seed(1000)
# Create data where Z1 is valid and Z2 is invalid
df <- tibble(Z1 = rnorm(1000), Z2 = rnorm(1000)) %>%
  mutate(X = Z1 + Z2 + rnorm(1000)) %>%
  # True effect is 1
  mutate(Y = X + Z2 + rnorm(1000))

iv <- feols(Y ~ 1 | X ~ Z1 + Z2, data = df, se = 'hetero')
fitstat(iv, 'sargan')
```
 - That's a small p-value! We can reject that the results are similar for each IV, telling us that one is endogenous (although without seeing the actual data generating process we couldn't guess if it were $Z1$ or $Z2$ )

---

# Overidentification tests: simulation

- How different are they? What did the test see that it was comparing? 

```{r, echo = TRUE}
iv1 <- feols(Y ~ 1 | X ~ Z1, data = df)
iv2 <- feols(Y ~ 1 | X ~ Z2, data = df)
export_summs(iv1, iv2, statistics = c(N = 'nobs'))
```

Notice the first model gives an accurate coefficient of 1

    
  